---
title: "Week 10"
author: "Penelope Pooler Eisenbies"
date: "`r Sys.Date()`"
output: 
  html_document: 
    code_folding: hide
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r setup, include=TRUE}

# this line specifies options for default options for all R Chunks
knitr::opts_chunk$set(echo=T, 
                      highlight=T)

# suppress scientific notation
options(scipen=100,
        getSymbols.warning4.0 = FALSE)

# install helper package (pacman), if needed
if (!require("pacman")) install.packages("pacman", repos = "http://lib.stat.cmu.edu/R/CRAN/")

# install and load required packages
# pacman should be first package in parentheses and then list others
pacman::p_load(pacman,tidyverse, ggthemes, magrittr, knitr, lubridate, gridExtra, RColorBrewer, 
               flexdashboard, maps, usdata, countrycode, mapproj, shadowtext, grid)

# verify packages (comment out in finished documents)
p_loaded()

```


---

### Upcoming Dates

- **Proposal were due on Monday, 10/31** 

  - If your group missed this deadline, you may still turn it in for 50% credit.

- Project presentations begin 5 weeks from today.

- HW 5 - Part 1 is due Wednesday, 11/2, at midnight (Grace period ends Friday 4/8 at midnight)

- HW 5 - Part 2 will be posted next week and due before Thanksgiving Break

<br>

- **Quiz 2 is on Thursday, 11/10**

  - Mostly on Weeks 5 through 9, but material is cumulative 

  - Mostly on HW Assignments 4 and 5, but material is cumulative
  
  - Data Mgmt. tasks will requires 2 or 3 steps and then you will answer questions 
  
---
  
### This Week and next Week
  
- Practice Questions for Quiz 2 will be posted Tomorrow (Wed. 11/2)
  
- Thu. 11/3: Skills/Concepts Review for Quiz 2

  - Putting skills together for different goals
  
  - Come with questions or submit them in Engagement Questions
  
<br>
  
- Today - Intro to Managing and Plotting Geographic Data 

  - Today's geographic maps will not be on Quiz 2.  
  
  - Mapping data is an effective data curation tool that may be useful in this class, other classes, your career.
  
<br>
  
- Tuesday, 11/8, we will review these skills and extend them to look at country data.

- **If you want help with project data, please reach out to me or TAs**.

---

### In-class Exercise 1 - Week 10

**Purpose:**  

- To gain some experience and understanding of map data available in R and elsewhere.

- To experiment with mapping data

- Students are encouraged to use domestic or international map data in their dashboards if appropriate.

---

### Plots by US County

#### Data Preparation

- The data for today's in-class exercise is part of R.  

- These geographic data are useful if you have information by state or county and you want to show a [choropleth map](https://en.wikipedia.org/wiki/Choropleth_map) of your data.

- R also has world information, e.g., countries, continents, etc.

```{r us data prep}
us_states <- map_data("state") |>                    # state polygons (not used today)
  rename("state" = "region")

us_counties <- map_data("county") |>                 # county polygons
  rename("state" = "region", "county" = "subregion")
# unique(us_counties$county[us_counties$state=="louisiana"])   # note issue Louisiana counties

cnty2019_all <- county_2019
# unique(cnty2019_all$county[cnty2019_all$state=="louisiana"]) # note issue Louisiana counties

cnty2019_all <- cnty2019_all |> 
  mutate(state = tolower(state),
         county = tolower(name),
         county = gsub(" county", "", county),
         county = gsub(" parish", "", county),
         county = gsub("\\.", "", county))            # \\ is required because . used in R coding

cnty2019_all <- full_join(us_counties,cnty2019_all)  # geo data and demographic data

```

---

### County Demographic Plots

- Creating a new dataset not required, but helpful
- Plot code could be converted into a function

```{r demographic example - select data and plots 1 thru 4}

# select data
cnty_data1 <- cnty2019_all |> 
  select(long:county, hs_grad, bachelors, 
         household_has_computer, 
         household_has_broadband) 


# hs grad pct plot
(cnty_hs_grad <- cnty_data1 |>
  ggplot(aes(x=long, y=lat, 
             group=group, fill=hs_grad)) +
    geom_polygon() +
    theme_map() +
    coord_map("albers", lat0 = 39, lat1 = 45) +
    labs(fill= "", 
         title="Percent with High School Degree") +
    scale_fill_continuous(type = "viridis"))


# bachelors pct plot
(cnty_bachelors <- cnty_data1 |>
  ggplot(aes(x=long, y=lat, 
             group=group, 
             fill=bachelors)) +
    geom_polygon() +
    theme_map() +
    coord_map("albers", lat0 = 39, lat1 = 45) +
    labs(fill= "", title="Percent with Bachelor's Degree") +
    scale_fill_continuous(type = "viridis"))


# hh has computer pct plot
(cnty_computer <- cnty_data1 |>
  ggplot(aes(x=long, y=lat, 
             group=group, 
             fill=household_has_computer)) +
    geom_polygon() +
    theme_map() +
    coord_map("albers", lat0 = 39, lat1 = 45) +
    labs(fill= "", title="Percent of Households with A Computer")+
    scale_fill_continuous(type = "viridis"))


# hh has broadband pct plot
(cnty_brdbd <- cnty_data1 |>
  ggplot(aes(x=long, y=lat, 
             group=group, 
             fill=household_has_broadband)) +
    geom_polygon() +
    theme_map() +
    coord_map("albers", lat0 = 39, lat1 = 45) +
    labs(fill= "", title="Percent of Households with BroadBand") +
    scale_fill_continuous(type = "viridis"))

```




---

### Demographic Plot Grid - Order Matters - L to R by Rows

```{r grid of all four demographic plots, fig.align='center', fig.dim=c(14,6), echo=F}

grid.arrange(cnty_hs_grad, cnty_computer, cnty_bachelors, cnty_brdbd, ncol=2) 

```

---

### In-class Exercise 1 - 2x2 Plot Grid - Steps 1 and 2

**Steps to Follow**

**`1.`** Examine the available variables in the **`cnty_2019_all`** dataset saved from R.
  
**`2.`** Create Individual Plots**.

- Variable names in R dataset and definitions:

   - **`household_has_smartphone`: Households with Smart Phones**  
   - **`median_age`: Median Age**
   - **`median_household_income`: Median Household Income**
   - **`median_individual_income`: Median Individual Income**
   
#### Create Individual Plots

You can copy provided plot code and modify it for these variables
or you can try to create a function (not required today).

```{r usmaps demographic maps exercise}

# select data for plot (not required, but helpful)
# cnty_data2 <- 

# create four individual plots, one for each variable
# use provided plot code and modify

```
   
---

### In-class Exercise 1 - 2x2 Plot Grid - Steps 3 and 4
  
**`3.`** Create 2x2 plot of these four variables by US County as part of your class participation credit for this week.

- **For full credit, plots must be in the order specified.**

   - **Row 1:** should have smartphone variable and median age.
   - **Row 2:** median household and individual income.

#### Create Plot Grid using `grid.arrange`

```{r  in-class exercise 2x2 plot grid}

# for full credit grid of plots must be in order specified
# border around all four plots not required
```

**`4.`** Export and submit plot grid

Reminder of two ways to export plots:

- Right click on plot, then save as...
- Use this R code, but modify for your plot and filename (height and width can also be modified)

```{r export code example, eval=F}

png("Wk2_Example_Plot_Grid_Penelope_Pooler.png", height=480, width=720)
grid.arrange(cnty_hs_grad, cnty_computer, cnty_bachelors, cnty_brdbd, ncol=2)
dev.off()

```

---

### Mapping Log Transformed Data

- The previous examples above are all percent data.

- No transformations are needed

- In contrast, population data or financial data are often right-skewed and need to be ((log transformed.**

- Recall from MAS 261, BUA 345 (and perhaps FIN classes):

  - An effective transformation for right skewed data is the natural log (LN) transformation.
  
  -  The following demo shows how useful it is for mapping right skewed data.


---

### Plot of Skewed Data

**If map looks like this, examine histogram of data.**

```{r untransformed pop map code}

cnty_data3 <- cnty2019_all |> 
  select(long:county, pop) |>
  mutate(pop1k = pop/1000)
       
(cnty_pop <- cnty_data3 |>
    ggplot(aes(x=long, y=lat, group=group, fill=pop1k)) +
    geom_polygon() +
    theme_map() +
    coord_map("albers", lat0 = 39, lat1 = 45) +
    labs(fill= "", title="Population by County",
         subtitle="Unit is 1000 People") +
    scale_fill_continuous(type = "viridis") +
    theme(legend.position = "bottom",
          legend.key.width = unit(1, "cm")))

```


---

### Histogram Clarifies Data Skewness

**Histogram confirms highly right-skewed data.**

```{r untransformed pop hist code}
(hist_pop <- cnty_data3 |>
  ggplot() +
  geom_histogram(aes(x=pop1k),
                 fill="lightblue", 
                 col="darkblue") +
  labs(x="Population", title="Histogram of US Population Data",
       y = "Count",subtitle="Unit is 1000 People") +
  theme_classic())
```

---

### Solution - Skewed Variable shown using Natural Log Scale

- Data are not transformed, but data axes and scale are
- Options specified in `scale_fill_continuous`:

  - `trans = "log"` 
  - `breaks = c(...)` 
  
- break intervals determined by examining data

<br>

**Transformed Data shown on log scale (see legend)**
  
```{r log transformed map plot code}

(cnty_lpop <- cnty_data3 |>
  ggplot(aes(x=long, y=lat, group=group, fill=pop1k)) +
    geom_polygon() +
    theme_map() +
    coord_map("albers", lat0 = 39, lat1 = 45) +
    labs(fill= "", title="Population by County",
         subtitle="Unit is 1000 People and Date are Log-transformed") +
    scale_fill_continuous(type = "viridis",trans="log",
                          breaks=c(1,10,100,1000,10000)) +
    theme(legend.position = "bottom",
          legend.key.width = unit(1, "cm")))

```


---

### Histogram of Log transformed Data

- In a final dashboard, usually don't show histograms
- Ideal for exploring the data.
- Scatterplots and boxplots are also useful for data exploration.

```{r log transformed pop hist code}

(hist_lpop <- cnty_data3 |>
  ggplot() +
  geom_histogram(aes(x=pop1k),fill="lightblue", col="darkblue") +
  labs(x="Population", 
       title="Histogram of Natural Log of US Population Data",
       y = "Count",
       subtitle="Unit is 1000 People and Data are Log-transformed") +
  theme_classic() + 
  scale_x_continuous(trans="log", breaks=c(1,10,100,1000,10000)))

```


---

### Population Plot Grid

```{r grid of all four population plots}
grid.arrange(cnty_pop, hist_pop, cnty_lpop, hist_lpop, ncol=2) 

```


---

### When and How to Log Transform

- Log transformation are useful if you have right skewed POSITIVE data such as

  - Prices
  - Population
  - Sales
  - Income
  
- Note: If data (x) have zeros, a good option is to use log(x + 1) 

  - ln(1) = 0 (In R `log(1)` = 0)
  
  - 0 values in the data will still be zeros
  
<br>
  
- In the following example we will create plots for number of households by county:

  - Histograms with and without LN transformation 
    
  - Map plots of with and without LN transformation 
  
---

### Number of Households Per County

#### Untransformed Data Histogram

```{r data and untransformed hh data histogram code}

cnty_data4 <- cnty2019_all |> 
  select(long:county, households) |>
  mutate(households1K = households/1000)

(hist_hholds <- cnty_data4 |>
  ggplot() +
  geom_histogram(aes(x=households1K),
                 fill="lightblue", 
                 col="darkblue") +
  labs(x="Number of Households", title="Histogram of US Household Data",
       y = "Count",subtitle="Unit is 1000 Households") +
  theme_classic())

```

---

### Number of Households Per County

#### Log transformed Data Histogram

```{r log transformed hh data histogram code}

(hist_lhholds <- cnty_data4 |>
  ggplot() +
  geom_histogram(aes(x=households1K),
                 fill="lightblue", 
                 col="darkblue") +
  labs(x="Number of Households", 
       title="Histogram of Natural Log of US Household Data",
       y = "Count",
       subtitle="Unit is 1000 Households and Data are Log-transformed") +
  theme_classic() + 
  scale_x_continuous(trans="log", breaks=c(1,10,100,1000)))

```


---

### Number of Households Per County

```{r untransformed hh plot map code}

(cnty_hholds <- cnty_data4 |>
  ggplot(aes(x=long, y=lat, group=group, fill=households1K)) +
  geom_polygon() +
  theme_map() +
  coord_map("albers", lat0 = 39, lat1 = 45) +
  labs(fill= "", title="Number of Households by County",
       subtitle="Unit is 1000 Households") +
  scale_fill_continuous(type = "viridis") +
  theme(legend.position = "bottom",
        legend.key.width = unit(1, "cm")))

```



---

**Log-transformed data map is more more informative about geographic variability.**

#### In-class Exercise 2: Submit R code to create log transformed households map in a text file with your name.**

```{r log transformed hh data map}

# cnty_lhholds <- cnty_data4 |>


```

---

### Households per County Plot Grid

```{r grid of all four household plots}

grid.arrange(hist_hholds, hist_lhholds, cnty_hholds, ncol=2)

```

---

### Quiz 2 Info 

- Questions and Material from Quiz 1 may be on Quiz 2

- Practice Questions will be posted by tomorrow night (Wed. 11/2)

  - Review Quiz 1 and Quiz 1 Practice Questions
  
  - Review Week HW 4 and HW 5 - Part 1 and recent lectures 
  
- Converting text (character) date information to a date using [lubridate](https://rawgit.com/rstudio/cheatsheets/main/lubridate.pdf) commands (Week 5)

  - Example R commands:`ymd`, `dmy`, `mdy`, `ym` combined with `paste` to combine columns
  
- Extracting year, month, or day from the date variable using lubridate commands (Week 6)

  - Example R commands: `year`, `month`, `quarter`, `wday`, `day`

- Converting an `xts` dataset to a tibble (standard R dataset) (Week 7)

  - Creating a lineplot from time series (non-xts) dataset

- Converting a tibble to an `xts` dataset

   + Creating an interactive hchart
   
<br>
   
- You should be familiar with the `bls_tidy` function we created and how to use it to import similar datasets.

- There will be datasets to be imported AND combined (joined)

  - Data sets can be joined by row.
  
  - You should know how to do the different joins we covered and what each one does:
  
    - `full_join`
    
    - `right_join`
    
    - `left_join`
    
    - `inner_join`
    
- Data sets can be stacked by column if the columns are identical
  
    - In BUA 455 we covered `bind_rows`
    
<br>

- Cleaning messy data (Week 5 and Weeks 8-9)

  - Dealing with text (character variables)
  
    - `gsub`
    
    - `separate`
    
    - `unite` or `paste` or `paste0`
    
    - `ifelse` can be be used for text OR for numeric data
    
    - `ifelse` followed by `factor` allows you to make any categorical variable you want.
    
- Other commands for modifying text:
 
   - `tolower` and `toupper`
   
   - `str_trim`, `str_squish` and `str_pad`
   
   - `gsub`
   
   - `ifelse`
   
   
---

### Additional Text Commands (Not on Quiz 2)

   - New commands (not for quiz): Capitalize 1st letter
   
   - [Commands to covert case](https://stringr.tidyverse.org/reference/case.html)
   
   - `str_to_title`: First letter of each word
     
   - `str_to_sentence`: First letter of first word
   
### Additional skills for Quiz 2 from HW 5 - Part 1

  - summing across rows using `sum(c_across(...))`
  
  - using `pivot_wider` and then `pivot_longer` and then replacing NAs with 0 to create a 'complete' data set
  
    - Useful for area plots
    
- Plotting skills for Quiz 2

   - line plot or area plot
   
   - hchart
   
   - possibly a grouped bar chart
   
---


### **Key Points from Week 10**

**Introduction to Geographic Data**

- Use skills already covered to

  - clean data and check text variables
  
  - join datasets
  
  - create plot grids for comparing variables 

- Determine if variables need to be log-transformed

- Quiz 2 Practice Questions will be Posted by 11/2


**You may submit an 'Engagement Question or Comment' about Week 10 lectures until Thursday, 11/3, at midnight on Blackboard.**

